An Innovative Storage Stack Addressing Extreme Scale Platforms and Big Data Applications

### 2014-04-17 paper.tex
fixed up the title and intro a bit. re-ordered the pieces Ivo added a bit to make it more sections by layer. More still needs to be done.

\title{An Innovative Storage Stack Addressing the Requirements of Extreme Scale Platforms and Big Data Applications}

@@

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve the needs of an extreme scale platform. To
address these limitations, a joint effort between the US Department of Energy's
Office of Advanced Simulation and Computing and Advanced Scientific Computing
Research commissioned an effort to develop a design and prototype for an IO
stack suitable for the extreme scale environment. This is a joint effort led by
Lawrence Livermore National Laboratory, with the DOE Data Management Nexus
leads Rob Ross and Gary Grider as coordinators and contract lead Mark Gary. The
participating labs are LLNL, SNL, LANL, ORNL, PNL, LBNL, and ANL. Additional
industrial partners contracted include the Intel Lustre team, EMC, DDN, and the
HDF Group. This team has developed a specification
set~\cite{fastforward:2014:docs} for a future IO stack to address the
identified challenges. The first phase recently completed with a second phase
currently getting underway as of early 2014. The core focus of the first phase
was basic functionality and design. Overriding many of the decisions during
this and any subsequent phases is the reality of budgets. Placing GBs of NVRAM
on every node, while a potentially advantageous approach, is not financially
feasible. With this in mind, the second phase will refine this design
incorporating fault recovery and other features missing from the first phase.

@@

The basic architecture incorporates five layers. The top layer is a high level
IO library, such as the demonstration HDF-5 library~\cite{hdf5}. The intent is
to only have access to the storage stack through such an API to manage the
complexity of working with the lower layers. Below the user API is an IO
forwarding layer that redirects IO calls from the compute nodes to the IO
dispatching layer.  This IO forwarding layer is analogous to the function of
the IO nodes in a BlueGene machine. The next two layers have considerable
functionality. The IO dispatcher (IOD) serves as the primary storage interface
for the IO stack and is the only way to reach the persistent storage array in
lower layers. The Distributed Application Object Storage (DAOS) layer serves as
the persistent storage interface and is intended to be the foundation on which
everything else is built with no dependence on any technologies specified above
it. For example, the IOD layer and burst buffers are not required for DAOS to
operate properly. At the bottom is the Versioning Object Storage Device (VOSD).
It serves as the interface for storing objects of all types efficiently.

%%%%%%%%%%%%%%%%%%%%%%%%% <<<<<<<<<<<<<<<<

The ``I/O nodes'' handle 
requests forwarded by the scientific applications, integrate a tier of
solid-state devices to absorb the burst of random operations, and
organize/re-format the data so that transfers from the staging area to
the traditional parallel file system can be done more efficiently. It
also has the capacity of executing analysis on the fresh data that has
just been generated by simulation applications running at the compute
nodes. As the data arrives, re-organization and data preparation can be
applied in order to anticipate the execution of analytical tasks
{[}3{]}.

@@

\subsection{HDF5 extensions}\label{hdf5-extensions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%% >>>>>>>>>>>>>>

The focus of this paper is primarily the IOD layer given the critical role it
has in the performance and functionality of the entire stack. Most of the key
features explored in this paper all have a strong presence in the IOD layer
motivating the focus of this examination.

The core idea for IOD is to provide a way to manage the IO load that is
separate from the compute nodes and the storage array. Communication intensive
activities, such as data rearrangement, can be moved to the IOD layer
offloading the communication load from the compute nodes. IOD has three main
purposes. First, if the optional burst buffer is available, it works as a fast
cache absorbing write operations for the slower trickle out to the central
storage array. It can also be used to retrieve objects from the central storage
array for more efficient read operations and offers data filtering to make
client reads more efficient.  Second, it offers the transaction mechanism for
controlling data set visibility and to manage faults that would prevent a data
set from being used. Third, data processing operations can be placed in the
IOD. These operations are intended to offer data rearrangement, filtering, and
similar operations prior to data reaching the central storage array.

While these ideas are not necessarily new, they are new twists on best of class
efforts for these technologies. For example, offloading the collective
two-phase data sieving from the compute nodes to reorganize data has proven
effective at reducing the total time for writing data due to fewer participants
involved in the communication patterns~\cite{lofstead:2011:nessie-staging}.
Beyond these broad items, there are many important details some of which are
examined in more detail below.

Along with the analysis of the published design documents, a discussion of the
design philosophy representing the overall intent is presented. This
information represents information that may or may not have been written down,
but is the intent of ultimate product. These insights were gained based on
personal conversations with the team
members~\cite{JohnBent,QuinceyKoziol,EricBarton} discussing some of the
potential challenges with the design as written. These ideas are presented to
give a fuller picture of where the project is going rather than dwelling on the
limitations of the published documents.

>>>>

\title{An Innovative Storage Stack Addressing Extreme Scale Platforms and Big Data Applications}

@@

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve the needs of an extreme scale platform.
Adding to the complexity of the problem is the variety of Big Data problems
that are taking advantage of these platforms. Trying to address both of these
domains is a daunting problem.

To address these challenges, a joint effort between the US Department of
Energy's Office of Advanced Simulation and Computing and Advanced Scientific
Computing Research commissioned an effort to develop a design and prototype for
an IO stack suitable for the extreme scale environment called the Fast Forward
Storage and IO (FFSIO) project. This is a joint effort led by Lawrence
Livermore National Laboratory, with the DOE Data Management Nexus leads Rob
Ross and Gary Grider as coordinators and contract lead Mark Gary. The
participating labs are LLNL, SNL, LANL, ORNL, PNL, LBNL, and ANL.  Additional
industrial partners contracted include the Intel Lustre team, EMC, DDN, and the
HDF Group. This team has developed a specification
set~\cite{fastforward:2014:docs} for a future IO stack to address the
identified challenges. The first phase recently completed with a second phase
currently getting underway as of early 2014. The core focus of the first phase
was basic functionality and design. Overriding many of the decisions during
this and any subsequent phases is the reality of budgets. Placing GBs of NVRAM
on every node, while a potentially advantageous approach, is not financially
feasible. With this in mind, the second phase will refine this design
incorporating fault recovery and other features missing from the first phase.

@@

Overall, the architecture shifts from the idea of files and directories to
containers of objects. This shift avoids the bottlenecks related to the POSIX
files and directories structure such as the serialization of file creation, the
limitation and impact of the number of files in a directory, and the limited
semantics of a byte stream. Instead, the new interface focuses on high-level
data models and their properties and relationships. This concept permeates the
entire IO stack.

In addition to addressing the traditional scientific workload, this project
seeks to expand functionality to better support Big Data type applications. The
key idea is to support Arbitrary Connected Graphs (ACGs) such as those used in
Map-Reduce systems. Key system features are introduced to efficiently support
these computing models in addition to the typically bursty IO loads of more
traditional HPC applications.

The basic architecture incorporates five layers some of which are potentially
optional. The top layer is a high level IO library, such as the demonstration
HDF-5 library~\cite{hdf5}. The intent is to only have access to the storage
stack through such an API to manage the complexity of working with the lower
layers. These changes incorporate the necessary features for ACGs.

Below the user API is an IO forwarding layer that redirects IO calls from the
compute nodes to the IO dispatching layer.  This IO forwarding layer is
analogous to the function of the IO nodes in a BlueGene machine. The next two
layers have considerable functionality. The IO dispatcher (IOD) serves as the
primary storage interface for the IO stack and is the only way to reach the
persistent storage array in lower layers. Ideally, the IOD layer can be
optional based on the available hardware. Much of the functionality offered at
this layer would shift either up or down the stack as discussed in detail
below. The Distributed Application Object
Storage (DAOS) layer serves as the persistent storage interface and is intended
to be the foundation on which everything else is built with no dependence on
any technologies specified above it. For example, the IOD layer and burst
buffers are not required for DAOS to operate properly. At the bottom is the
Versioning Object Storage Device (VOSD).  It serves as the interface for
storing objects of all types efficiently.

%The core idea for IOD is to provide a way to manage the IO load that is
%separate from the compute nodes and the storage array. Communication intensive
%activities, such as data rearrangement, can be moved to the IOD layer
%offloading the communication load from the compute nodes. IOD has three main
%purposes. First, if the optional burst buffer is available, it works as a fast
%cache absorbing write operations for the slower trickle out to the central
%storage array. It can also be used to retrieve objects from the central storage
%array for more efficient read operations and offers data filtering to make
%client reads more efficient.  Second, it offers the transaction mechanism for
%controlling data set visibility and to manage faults that would prevent a data
%set from being used. Third, data processing operations can be placed in the
%IOD. These operations are intended to offer data rearrangement, filtering, and
%similar operations prior to data reaching the central storage array.

%While these ideas are not necessarily new, they are new twists on best of class
%efforts for these technologies. For example, offloading the collective
%two-phase data sieving from the compute nodes to reorganize data has proven
%effective at reducing the total time for writing data due to fewer participants
%involved in the communication patterns~\cite{lofstead:2011:nessie-staging}.
%Beyond these broad items, there are many important details some of which are
%examined in more detail below.

Along with the analysis of the published design documents, a discussion of the
design philosophy representing the overall intent is presented. This
information represents information that may or may not have been written down,
but is the intent of ultimate product. These insights were gained based on
personal conversations with the team
members~\cite{JohnBent,QuinceyKoziol,EricBarton} understanding the current
design as written. These ideas are presented to
give a fuller picture of where the project is going rather than dwelling on the
limitations of the published documents.

\section{HDF Layer}
The API has been maintained as much as possible to make backwards compatibility
easier. Instead, the Virtual Object Layer (VOL), a new layer under development
for several years has been leveraged to incorporate the needed support.

Some additional API functions have been introduced to support....

\section{I/O Nodes}
The ``I/O nodes'' handle 
requests forwarded by the scientific applications, integrate a tier of
solid-state devices to absorb the burst of random operations, and
organize/re-format the data so that transfers from the staging area to
the traditional parallel file system can be done more efficiently. It
also has the capacity of executing analysis on the fresh data that has
just been generated by simulation applications running at the compute
nodes. As the data arrives, re-organization and data preparation can be
applied in order to anticipate the execution of analytical tasks
{[}3{]}.

@@

@@
